---
title: "2.2 Variability"
author: "Dr. Lauren Perry"
output:
  beamer_presentation: default
  ioslides_presentation:
    widescreen: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Goals

1. Calculate and interpret measures of variation/spread.
    - standard deviation and interquartile range
    - determine which measure of variation to use for a given dataset
2. Find and interpret measures of position.
    - minimum and maximum
    - quartiles
3. Summarize data using box plots.

## Measures of Variability

How much do the data vary?

Should we care? Yes! The more variable the data, the harder it is to be confident in our measures of center!

- If you live in a place with extremely variable weather, it is going to be much harder to be confident in how to dress for tomorrow's weather. 
- If you live in a place where the weather is always the same, it's much easier to be confident in what you plan to wear.

## Range

One easy way to think about variability is the **range** of the data: $$\text{range} = \text{maximum} - \text{minimum}$$ 

- This is quick and convenient, but it is *extremely* sensitive to extreme values! 
- It also takes into account only two of the observations.
    - We would prefer a measure of variability that takes into account *all* the observations. 

## Deviation

**Deviation** is the distance of an observation from the mean: $$x - \bar{x}$$

- We want to think about how far - on average - a typical observation is from the center.
- Might think to take the average deviance... but it turns out that summing up the deviances will *always* result in 0! 
    - Conceptually, this is because the stuff below the mean (negative numbers) and the stuff above the mean (positive numbers) end up canceling each other out until we end up at 0.
    
***

One way to deal with this is to make all of the numbers positive, which we accomplish by *squaring* the deviance.

|     | Deviance | Squared Deviance |
| :-: | :-: | :-: |
| $x$ | $x - \bar{x}$ | $(x - \bar{x})^2$ |
| 2 | -1.2 | 1.44 |
| 5 | 1.8 | 3.24 |
| 3 | -0.2 | 0.04 |
| 4 | 0.8 | 0.64 |
| 2 | -1.2 | 1.44 |
| $\bar{x}=3.2$ | Total = 0 | Total = 6.8 |

## Variance

**Variance** (denoted $s^2$) is the "average" squared distance from the mean:
\[
  s^2 = \frac{(x_1-\bar{x})^2 + (x_2-\bar{x})^2 + \dots + (x_n-\bar{x})^2}{n-1}
\]
where $n$ is the sample size. 

- Notice that we divide by $n-1$ and NOT by $n$. 

## Standard Deviation

Finally, we come to **standard deviation** (denoted $s$). 

- The standard deviation is the square root of the variance. $$s = \sqrt{s^2}$$
- We will use a computer to calculate the variance and standard deviation.

Interpretation: A "typical" observation is within about one standard deviation of the mean (between $\bar{x}-s$ and $\bar{x}+s$).

## Interquartile Range

The **interquartile range** (**IQR**) represents the middle 50% of the data. 

- This is another measure of variability!

## Percentiles

- Recall that the *median* cut the data in half: 50% of the data is below and 50% is above the median. 
- This is also called the **50th percentile**. 
- The **$p$th percentile** is the value for which $p$% of the data is below it.

***

To get the middle 50%, we will split the data into four parts:

| 1	| 2	| 3	| 4 |
| :-: | :-: | :-: | :-: |
| 25%	| 25%	| 25%	| 25% |

The 25th and 75th percentiles, along with the median, divide the data into four parts. 

## Quartiles

We call these three measurements the **quartiles**:

- **Q1**, the first quartile, is the median of the lower 50% of the data.
- **Q2**, the second quartile, is the median.
- **Q3**, the third quartile, is the median of the upper 50% of the data.

## Example

Consider {1, 2, 3, 4, 5, 6, 7, 8, 9, 10}

- Cutting the data in half: {1, 2, 3, 4, 5 | 6, 7, 8, 9, 10}
- So the median (Q2) is $\frac{5+6}{2}=5.5$.
- Q1 is the median of {1, 2, 3, 4, 5}, or 3
- Q3 is the median of {6, 7, 8, 9, 10}, or 8

**Note:** this is a "quick and dirty" way of finding quartiles. A computer will give a more exact result. 

## Interquartile Range

Then the interquartile range is
\[
  \text{IQR} = \text{Q3}-\text{Q1}
\]

- The IQR is resistant to extreme values. 

## Which measure to use?

- mean and standard deviation when the data are symmetric 
- median and IQR when the data are skewed

## Box Plots

Box plots summarize the data with 5 statistics plus extreme observations:

```{r}
set.seed(0)
x <- c(rpois(45,2),-2.5)
boxplot(x, ylim=c(-3,7), col="lightblue")
legend(x=0.62,y=6.8,legend="Upper Whisker",box.lty=0)
legend(x=0.68,y=3.8,legend="Q3",box.lty=0)
legend(x=1.2,y=2.8,legend="Median",box.lty=0)
legend(x=0.68,y=1.8,legend="Q1",box.lty=0)
legend(x=1.1,y=0.8,legend="Lower Whisker",box.lty=0)
legend(x=0.68,y=-1.7,legend="Potential Outlier",box.lty=0)
```

## Drawing a box plot:

1. Draw the vertical axis to include all possible values in the data.
2. Draw a horizontal line at the median, at $\text{Q1}$, and at $\text{Q3}$. Use these to form a box.
3. Draw the **whiskers**. The whiskers' upper limit is $\text{Q3}+1.5\times\text{IQR}$ and the lower limit is $\text{Q1}-1.5\times\text{IQR}$. The actual whiskers are then drawn *at the next closest data points within the limits*.
5. Any points outside the whisker limits are included as individual points. These are **potential outliers**. 

We won't draw box plots by hand, but understanding how they are drawn will help us understand how to interpret them!

## Outliers

(Potential) outliers can help us...

- examine skew (outliers in the negative direction suggest left skew; outliers in the positive direction suggest right skew).
- identify issues with data collection or entry, especially if the value of the outliers doesn't make sense.

## Descriptive Measures for Populations

- We've thought about calculating various descriptive statistics from a sample.
- Our long-term goal is to estimate descriptive information about a population. 
- At the population level, these values are called **parameters**. 

***

- When we find a measure of center, spread, or position, we use a sample to calculate a single value. 
- These single values are called **point estimates**.
    - They are used to *estimate* the corresponding population parameter. 

| Point Estimate | Parameter |
|-|-|
| sample mean: $\bar{x}$ | population mean: $\mu$ |
| sample standard deviation: $s$ | population standard deviation: $\sigma$ |

