---
title: "3.1 Linear Equations"
author: "Dr. Lauren Perry"
output:
  ioslides_presentation: 
    widescreen: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Goals

1. Review linear equations.
2. Motivate regression.
3. Interpret a slope and intercept in context. 
4. Use a regression line to predict values of a dependent variable.

## Linear Equations

- Should already have seen linear equations like $$y=mx+b$$
- In statistics, we write these as
$$y=b_0 + b_1x$$ 
    - $b_0$ and $b_1$ are constants.
    - $x$ is the **independent variable**.
    - $y$ is the **dependent variable** 

## Slope and Intercept

$$y=b_0 + b_1x$$ 

- The **y-intercept** is $b_0$, the value of $y$ when $x=0$. 
- The **slope** is $b_1$, the change in $y$ for a 1-unit change in $x$.

***

A **scatterplot** shows the relationship between two (numeric) variables.

```{r}
plot(1000*mtcars$wt, mtcars$mpg, pch=16,
     main = "Scatterplot of Car Weight vs MPG",
     xlab="Weight (lbs)", ylab="Miles Per Gallon")
```

We call this type of data **bivariate data**.

***

```{r}
set.seed(0)
num <- sample(1:30, 10)
price <- 8 + 3.25*num
plot(num,price,xlab="Number of Coffees Purchased in a Month", ylab="Amount of Money Spent on Coffee",
     cex=1.5, pch=16)
abline(a=8,b=3.25,lwd=1.5)
```

This relationship can be modeled perfectly with a straight line: $$y = 8 + 3.25x$$ 

## Example: Interpret Slope and Intercept

$$y = 8 + 3.25x$$ where $x$ is the number of coffees purchased in a month and $y$ is the amount of money spent on coffee.

- The *intercept* is 8, which is the dollar amount of money spent on coffee (the value of $y$) when 0 coffees are purchased in that month (when $x=0$).
- The *slope* is 3.25, which is the increase in amount of money spent on coffee (increase in $y$) for each additional coffee purchased (a one-unit increase in $x$).

***

- But what if that pound of coffee didn't always cost \$8? 
- Or the coffee drinks didn't always cost \$3.25? 

```{r}
set.seed(0)
num <- sample(1:30, 10)
price <- 8 + 3.25*num + rnorm(10,0,15)
plot(num,price,xlab="Number of Coffees Purchased in a Month", ylab="Amount of Money Spent on Coffee",
     cex=1.5, pch=16)
```

***

The linear regression line looks like $$y = \beta_0 + \beta_1x + \epsilon$$

-   $\beta$ is the Greek letter "beta".
-   $\beta_0$ and $\beta_1$ are constants.
-   Error (the fact that the points don't all line up perfectly) is
    represented by $\epsilon$.

***

We estimate $\beta_0$ and $\beta_1$ using data and denote the estimated line by $$\hat{y} = b_0 + b_1x $$

-   $\hat{y}$, "y-hat", is the estimated value of $y$.
-   $b_0$ is the estimate for $\beta_0$.
-   $b_1$ is the estimate for $\beta_1$.
-   ...and 0 is the estimate for $\epsilon$ (so we ignore it).

***

We use a regression line to make predictions about $y$ using values of
$x$. 

Think of this as the 2-dimensional version of a point estimate!

-   $y$ is the **response variable**.
-   $x$ is the **predictor variable**.

## Example

*Example*: Researchers took a variety of measurements on 344 adult
penguins in Antarctica.

```{r}
library(palmerpenguins)
#head(penguins)
plot(penguins$flipper_length_mm, penguins$body_mass_g,
     xlab = "Flipper Length (mm)", ylab = "Body Mass (g)")
# lm(penguins$body_mass_g ~ penguins$flipper_length_mm)
```

## Example

The regression model for these data is $$\hat{y}=-5780.83 + 49.69x$$

```{r}
temp <- summary(lm(penguins$body_mass_g ~ penguins$flipper_length_mm))
plot(penguins$flipper_length_mm, penguins$body_mass_g,
     xlab = "Flipper Length (mm)", ylab = "Body Mass (g)")
abline(temp$coefficients[1,1], temp$coefficients[2,1], lwd=2, col="red")
```

## Example

To predict the body mass for a penguin with a flipper length of 180mm, we just need to plug in 180 for flipper length ($x$):
$$\hat{y}=-5780.83 + 49.69\times 180 = 3163.37\text{g}.$$ 
- Note that the regression line automatically deals with units.

## Homework Problems

Section 3.1 Exercises 1 and 2