---
title: "3.2 Correlation"
author: "Dr. Lauren Perry"
output:
  beamer_presentation: default
  ioslides_presentation:
    widescreen: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Goals 

1.  Interpret a correlation coefficient.
    - Describe the strength of a linear relationship. 
    - Relate correlation coefficients to scatter plots. 

## Correlation

Goal: formalize the concept of the *strength* of a linear relationship.

The **correlation** (or **correlation coefficient**) $R$ between two variables describes the strength of their linear relationship.

$$R = \frac{1}{n-1}\sum_{i=1}^n\left(\frac{x_i - \bar{x}}{s_x}\times\frac{y_i - \bar{y}}{s_y}\right)$$

- $s_x$ and $s_y$ are the respective standard deviations for $x$ and
$y$
- The sample size $n$ is the total number of $(x,y)$ pairs.
- $R$ always takes values between $-1$ and $1$. 

This is a pretty involved formula! We'll let a computer handle this one.

***

Correlations

-   close to $-1$ suggest strong, negative linear relationships.
-   close to $+1$ suggest strong, positive linear relationships.
-   close to $0$ have little-to-no linear relationship.

*** 

Note: the sign of the correlation will match the sign of the slope!

-   If $R < 0$, there is a downward trend and $b_1 < 0$.
-   If $R > 0$, there is an upward trend and $b_1 > 0$.
-   If $R \approx 0$, there is no relationship and $b_1 \approx 0$.

## Examples

```{r}
set.seed(0)
par(mfrow=c(1,2))
x1 <- 1:100 + rnorm(100, 0, 25)
y1 <- 7*x1 + 4 + rnorm(100, 0, 100)
plot(x1, y1, main = "R = 0.93", xlab="", ylab="", xaxt="n", yaxt="n")

x2 <- 1:100 + rnorm(100, 0, 35)
y2 <- 5*x2 - 6 + rnorm(100, 0, 300)
plot(x2, y2, main = "R = 0.55", xlab="", ylab="", xaxt="n", yaxt="n")
```

## More Examples

```{r}
set.seed(0)
par(mfrow=c(1,2))
x3 <- 1:100 + rnorm(100, 0, 5)
y3 <- runif(100, 10, 100)
plot(x3, y3, main = "R = 0.07", xlab="", ylab="", xaxt="n", yaxt="n")

x4 <- 1:100 + rnorm(100, 0, 5)
y4 <- -runif(100, 5, 50) + rnorm(100, 0, 200)
plot(x4, y4, main = "R = -0.05", xlab="", ylab="", xaxt="n", yaxt="n")
```

## Even More Examples

```{r}
set.seed(0)
par(mfrow=c(1,2))
x5 <- 1:100 + rnorm(100, 0, 5)
y5 <- -5*x2 - 6 + rnorm(100, 0, 170)
plot(x5, y5, main = "R = -0.53", xlab="", ylab="", xaxt="n", yaxt="n")

x6 <- 1:100 + rnorm(100, 0, 5)
y6 <- -8*x6 + 15 + rnorm(100, 0, 100)
plot(x6, y6, main = "R = -0.94", xlab="", ylab="", xaxt="n", yaxt="n")
```

***

When two variables are highly correlated ($R$ close to $-1$ or $1$)

- we know there is a strong *relationship* between them
- BUT we do not know what *causes* that relationship

That is, **correlation does not imply causation**.

***
A final note:

- Obviously there's a strong relationship between $x$ and $y$. 
  - In fact, $y = x^2$. 
- But the *correlation* between $x$ and $y$ is 0!

```{r}
x <- -10:10
y <- x^2
plot(x,y, type='l', lwd=2); points(x,y, cex = 2, pch=16, col='salmon')
```
