---
title: "3.4 Finding a Regression Line"
author: "Dr. Lauren Perry"
output:
  ioslides_presentation: 
    widescreen: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Residuals

**Residuals** are the leftover *stuff* (variation) in the data after
accounting for model fit:
$$\text{data} = \text{prediction} + \text{residual}$$ 

## Residuals

- Each observation has its own residual. 
- The residual for an observation $(x,y)$ is the
difference between observed ($y$) and predicted ($\hat{y}$):
$$e = y - \hat{y}$$ 
- We denote the residuals by $e$ and find $\hat{y}$ by
plugging $x$ into the regression equation. 

Note: If an observation lands above
the regression line, $e > 0$. If below, $e < 0$.

## Residuals

Goal: get each residual as close to 0 as possible. 

To shrink the residuals toward 0, we minimize:
\begin{align}
\sum_{i=1}^n e_i^2 &= \sum_{i=1}^n (y_i - \hat{y}_i)^2 \\
& = \sum_{i=1}^n [y_i - (b_0 + b_1 x_i)]^2
\end{align} 

The values $b_0$ and $b_1$ that minimize this will make up our regression line.

## Finding $b_0$ and $b_1$

- The slope is
$$b_1 = \frac{s_y}{s_x}\times R$$ 
- The intercept is
$$b_0 = \bar{y} - b_1 \bar{x}$$

## Example: Old Faithful Geyser in Yellowstone

- `eruptions`, the length of each eruption
- `waiting`, the time between eruptions

```{r}
plot(faithful$waiting, faithful$eruptions, xlab="Waiting Time", ylab="Eruption Duration")
```

## Example: Old Faithful Geyser in Yellowstone

The sample statistics for these data are

 |      |    `waiting`    |  `eruptions`   |
 |:----:|:---------------:|:--------------:|
 | mean | $\bar{x}=70.90$ | $\bar{y}=3.49$ |
 |  sd  |   $s_x=13.60$   |   $s_y=1.14$   |
 |      |                 |   $R = 0.90$   |
 
 Find the regression line and interpret the parameters.
 
## Example: Old Faithful Geyser in Yellowstone

```{r, message=F}
plot(x = faithful$waiting, y = faithful$eruptions,
     xlab = "Waiting Time (minutes)",
     ylab = "Eruption Duration (minutes)")
abline(summary(lm(formula = eruptions ~ waiting, data=faithful))$coef[,1],
       col = "red", lwd = 2)
```

## Checkpoint

Consider a dataset on `height` and `age` of $n=84$ Loblolly pine trees. 

- We want to use tree `height` to predict tree `age`.

The sample statistics for these data are

 |      |    `height`     |      `age`     |
 |:----:|:---------------:|:--------------:|
 | mean | $32.36$ | $13.00$ |
 |  sd  |   $20.67$   |   $7.90$   |
 |      |                 |   $R = 0.99$   |
 
 Find the regression line and interpret the parameters.

## The Coefficient of Determination

The **coefficient of determination**, $R^2$, is the square of the correlation coefficient.

- This value tells us how much of the variability around the regression line is accounted for by the regression. 
- An easy way to interpret this value is to assign it a letter grade. 
    - For example, if $R^2 = 0.84$, the predictive capabilities of the regression line get a B.
